Discovering governing reactions from concentration data

### 2017-11-02 docs/manuscript.tex
wip: manuscript

	\section{Introduction}
	When presented with a non-equilibrium time series of possibly noisy concentration fluctuations of some species as output of, e.g., measurements from experiments or simulations that were parameterized by microscopic rates \textcolor{red}{(cite ReaDDy?)}, one can ask for the corresponding macroscopic rates and generating reaction network. 
	In this paper we present an application of the shallow learning method SINDy \cite{Brunton2015}. It is able to identify the generating parsimonious nonlinear dynamics in data that stems from dynamical systems, thus providing an interpretable result.
	In our application we, opposed to the original method, do not only look for macroscopic rates of net species change but investigate the specific reactions that might have lead to the observations.
	We demonstrate the algorithm on two toy problems - one problem showing that when there is no ambiguity in the system, one converges to the correct rates with increasing resolution of concentration fluctuations and one in which we compute a sparse reaction network for given data.

	\section{The method}
	The underlying model that we want to fit the data to is a law of mass action type dynamical system. To this end, let $S$ be the number of species, then the concentration data at a time $t$ can be represented by a vector
	\begin{align}
	\mathbf{x}(t)=\begin{pmatrix}
	x_1(t)\\ \vdots \\ x_S(t)
	\end{pmatrix}\in \mathbb{R}^S.
	\end{align}
	Further, one can choose $R$ possible ansatz reactions with their respective reaction function
	\begin{align}
	\textbf{y}_r(\textbf{x}(t))=\begin{pmatrix}
	y_{r,1}(\textbf{x}(t)) \\ \vdots \\ y_{r,S}(\textbf{x}(t))
	\end{pmatrix}
	\end{align}
	so that the change of concentration for species $i$ at time $t_i$, is represented by the dynamical system
	\begin{align}
	\dot{\textbf{x}}_i(t) = \sum_{r=1}^{R}y_{r,i}(\textbf{x}(t))\xi_r,\quad i=1,\ldots, S,
	\label{method:the-system}\end{align}
	where $\xi_r$ are the to-be estimated macroscopic rates.

	When presented with a time series with $T$ observations, the data can be represented as a matrix
	\begin{align}
	\textbf{X} = \begin{pmatrix}
		x_1(t_1) & x_2(t_1) & \cdots & x_S(t_1) \\
		x_1(t_2) & x_2(t_2) & \cdots & x_S(t_2) \\
		\vdots   & \vdots   & \ddots & \vdots   \\
		x_1(t_T) & x_2(t_T) & \cdots & x_S(t_T)
	\end{pmatrix} \in \mathbb{R}^{T\times S}.
	\end{align}
	Given this matrix, one can propose a library $\Theta(\textbf{X}) = \begin{pmatrix} \theta_1(\textbf{X}) & \theta_2(\textbf{X}) & \cdots & \theta_R(\textbf{X}) \end{pmatrix}$ of $R$ ansatz reactions with corresponding reaction functions
	\begin{align}
		\theta_r(\textbf{X}) = \begin{pmatrix}
		\textbf{y}_r(\textbf{X}_1)^T \\ \vdots \\ \textbf{y}_r(\textbf{X}_T)^T
		\end{pmatrix}\in \mathbb{R}^{T\times S},\quad r=1,\ldots,R,
	\label{method:the-reactions}\end{align}
	where $\textbf{X}_i$ denotes the $i$-th row in $X$. Applying the concentration trajectory to the library yields a data tensor $\Theta(\textbf{X})\in\mathbb{R}^{T\times S\times R}$. Following the approach of SINDy, the goal is now to find coefficients $\Xi = \begin{pmatrix} \xi_1 & \xi_2 & \cdots & \xi_R
	\end{pmatrix}^T$, so that
	\begin{align}
	\dot{\textbf{X}} = \Theta(\textbf{X})\Xi = \sum_{r=1}^{R}\theta_r(\textbf{X})\xi_r.
	\end{align}
	In particular, the system is linear in the coefficients $\Xi$, which makes sparse regression tools such as LASSO \cite{Tibshirani1996, Hastie2009} applicable. To this end, one can consider the minimization problem to find $\hat{\Xi}$ such that
	\begin{align}
		\hat{\Xi} = \underset{\Xi}{\arg\min}\left( \frac{1}{2T}\left\| \dot{\textbf{X}} - \Theta(\textbf{X})\Xi \right\|_F^2 + \alpha\|\Xi\|_1 \right) \quad \text{subject to }\Xi \geq 0,
	\label{method:minimizationproblem}\end{align}
	where $\alpha\geq 0$ is a sparsity inducing hyperparameter and $\|\cdot\|_F$ denotes the Frobenius norm. For $\alpha=0$ this problem reduces to constrained least-squares. For solving (\ref{method:minimizationproblem}) we apply the sequential least squares minimizer SLSQP, originally described in \cite{Kraft1988}, contained in, e.g., the software package SciPy \cite{SciPy}. As initial guess for the regulated problem we use the solution of (\ref{method:minimizationproblem}) with $\alpha=0$.

@@

	For generating time series data of concentrations we use the Gillespie method \cite{Gillespie1976}. For a given set of initial conditions, we produce several realizations which then are converted to a trajectory with fixed time step and averaged. Since these trajectories are piecewise constant, one has to take special care when approximating the temporal derivative. Simply taking finite differences has the effect that, with decreasing time step, the derivative is most of the time constant zero and approaches infinity at the jump discontinuities. To counter this effect, we beforehand perform a linear approximation between each two adjacent discontinuities, as depicted in Figure \ref{examples:linear-approximation}. (\textcolor{red}{Use regularized derivative?})

@@

	Here we demonstrate that, with decreasing time step $\Delta t$ and $\alpha=0$, the estimated rates converge to the true rates if there is no ambiguity in the system. The systems contains two species $A$ and $B$ which can convert into each other with macroscopic rates of $r_1=4.0$ and $r_2=0.5$, respectively. The initial amount of particles is set to $70$ of type $A$ and $0$ of type $B$. A possibly resulting change of concentrations over time is depicted in Figure \ref{ex1:concentration-curves}.

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/aux_ex1_concentration_curves}
		\end{center}
		\caption{A realization of the example given in Section \ref{sec:ex1}.}
		\label{ex1:concentration-curves}
	\end{figure}

	For each $\Delta t$, we averaged eight different realizations with identical starting conditions into one single trajectory and estimated the rates for reactions $A\to B$ and $B\to A$. This procedure, we repeated $30$ times. In Figure \ref{ex1:rates-convergence}, the average estimates alongside with their standard deviations are shown. One can see that, with decreasing $\Delta t$, also the estimate for the rates becomes better and they can be recovered.

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/aux_rates_convergence}
		\end{center}
		\caption{Convergence of estimated rates to the ground truth with decreasing $\Delta t$.}
		\label{ex1:rates-convergence}
	\end{figure}

@@

	The problem now is to find a suitable $\alpha$ such that a parsimonious solution is picked. To this end, cross-validation is applied. The trajectories that were used as training and test data set are depicted in Figure \ref{ex2:test-train-concentration-curves}. \textcolor{red}{Perhaps use several test trajectories with different initial conditions to get something like a confidence interval?} As an indicator to the estimation error we use the scoring function
	\begin{align}
		f_\alpha = \frac{1}{T}\|\mathbf{\dot{X}}_\text{test} - \Theta (\textbf{X}_\text{test}) \Xi_\text{train}\|_F^2.
		\label{ex2:score-function}
	\end{align}
	For a selection of values $\alpha\in [0, 100]$, the model parameters $\Xi$ were estimated and the scoring function (\ref{ex2:score-function}) evaluated, yielding the graph that is depicted in Figure \ref{ex2:score-graph}.

@@

	One can see that the minimum is at about
	\begin{align}
		\hat{\alpha} = \underset{\alpha\in[0,100]}{\arg\min}f_\alpha\approx 10.
	\end{align}
	When evaluating the minimization problem (\ref{method:minimizationproblem}) with $\hat{\alpha}$ as hyperparameter for the regularizer and applying a cut-off of $10^{-10}$ to the yielded parameters, one obtains the simpler set of reactions $A\xrightarrow{\hat{k}_1}D$, $D\xrightarrow{\hat{k}_3} B$, and $A+B\xrightarrow{\hat{k}_4}C$ with rates $\hat{k}_1 \approx 3.5\cdot 10^{-3}$, $\hat{k}_3\approx 9.1\cdot 10^{-3}$, and $\hat{k}_4\approx 8.7\cdot 10^{-5}$, respectively.

>>>>

	\section{Introduction}
	When presented with a time series of possibly noisy non-equilibrium concentration fluctuations of some species as output of, e.g., measurements from experiments or simulations that were parameterized by microscopic rates \textcolor{red}{(cite ReaDDy?)}, one can ask for the corresponding macroscopic rates and a generating reaction network. 
	In this paper we present an application of the shallow learning method SINDy \cite{Brunton2015}. By sparse regression, it is able to identify generating nonlinear dynamics in data that stems from dynamical systems. The parsimonious nature of the results avoids overfitting and provides interpretability.
	In our application we, as opposed to the original method, do not only look for macroscopic rates of net species change but investigate the specific reactions that might have lead to the observations.
	We demonstrate the algorithm on two toy problems - one problem showing that when there is no ambiguity in the system, one converges to the correct rates with increasing resolution of concentration fluctuations and one in which we compute a sparse reaction network for given data.

	\section{The method}
	The underlying model is a law of mass action type dynamical system. To this end, let $S$ be the number of species, then the observed concentration at a time $t$ can be represented by a vector
	\begin{align}
	\mathbf{x}(t)=\begin{pmatrix}
	x_1(t)\\ \vdots \\ x_S(t)
	\end{pmatrix}\in \mathbb{R}^S.
	\end{align}
	Further, one can choose $R$ possible ansatz reactions with their respective reaction function
	\begin{align}
	\textbf{y}_r(\textbf{x}(t))=\begin{pmatrix}
	y_{r,1}(\textbf{x}(t)) \\ \vdots \\ y_{r,S}(\textbf{x}(t))
	\end{pmatrix}
	\end{align}
	so that the change of concentration for species $i$ at time $t$, is represented by the dynamical system
	\begin{align}
	\dot{\textbf{x}}_i(t) = \sum_{r=1}^{R}y_{r,i}(\textbf{x}(t))\xi_r,\quad i=1,\ldots, S,
	\label{method:the-system}\end{align}
	where $\xi_r$ are the to-be estimated macroscopic rates.

	When presented with a time series consisting of $T$ observations, the data can be represented as a matrix
	\begin{align}
	\textbf{X} = \begin{pmatrix}
		x_1(t_1) & x_2(t_1) & \cdots & x_S(t_1) \\
		x_1(t_2) & x_2(t_2) & \cdots & x_S(t_2) \\
		\vdots   & \vdots   & \ddots & \vdots   \\
		x_1(t_T) & x_2(t_T) & \cdots & x_S(t_T)
	\end{pmatrix} \in \mathbb{R}^{T\times S}.
	\end{align}
	Given this matrix, one can propose a library $\Theta(\textbf{X}) = \begin{pmatrix} \theta_1(\textbf{X}) & \theta_2(\textbf{X}) & \cdots & \theta_R(\textbf{X}) \end{pmatrix}$ of $R$ ansatz reactions with corresponding reaction functions
	\begin{align}
		\theta_r(\textbf{X}) = \begin{pmatrix}
		\textbf{y}_r(\textbf{X}_1)^T \\ \vdots \\ \textbf{y}_r(\textbf{X}_T)^T
		\end{pmatrix}\in \mathbb{R}^{T\times S},\quad r=1,\ldots,R,
	\label{method:the-reactions}\end{align}
	where $\textbf{X}_i$ denotes the $i$-th row in $X$. Applying the concentration trajectory to the library yields $\Theta(\textbf{X})\in\mathbb{R}^{T\times S\times R}$. Following the approach of SINDy, the goal is now to find coefficients $\Xi = \begin{pmatrix} \xi_1 & \xi_2 & \cdots & \xi_R
	\end{pmatrix}^T$, so that
	\begin{align}
	\dot{\textbf{X}} = \Theta(\textbf{X})\Xi = \sum_{r=1}^{R}\theta_r(\textbf{X})\xi_r.
	\end{align}
	In particular, the system is linear in the coefficients $\Xi$, which makes sparse regression tools such as LASSO \cite{Tibshirani1996, Hastie2009} applicable. To this end, one can consider the minimization problem to find $\hat{\Xi}$ such that
	\begin{align}
		\hat{\Xi} = \underset{\Xi}{\arg\min}\left( \frac{1}{2T}\left\| \dot{\textbf{X}} - \Theta(\textbf{X})\Xi \right\|_F^2 + \alpha\|\Xi\|_1 \right) \quad \text{subject to }\Xi \geq 0,
	\label{method:minimizationproblem}\end{align}
	where $\|\cdot\|_F$ denotes the Frobenius norm and $\alpha\geq 0$ is a sparsity inducing hyperparameter. For $\alpha=0$ the minimization problem reduces to constrained least-squares. In order to solve (\ref{method:minimizationproblem}) we apply the sequential least-squares minimizer SLSQP, originally described in \cite{Kraft1988}, by the software package SciPy \cite{SciPy}. As initial guess for the regularized problem we use the solution of (\ref{method:minimizationproblem}) with $\alpha=0$.

@@

	For generating time series data of concentrations we use the Gillespie method \cite{Gillespie1976}. For given initial conditions, we produce several realizations which then are converted to a trajectory with fixed time step and averaged. Since these trajectories are piecewise constant, one has to take special care when approximating the data's temporal derivative that is needed in (\ref{method:minimizationproblem}). Applying finite differences has the effect that, with decreasing time step, the derivative mostly constant zero and approaches infinity at the jump discontinuities. To counter this effect, we beforehand perform a linear approximation between each two adjacent discontinuities, as depicted in Figure \ref{examples:linear-approximation} and thus obtain a piecewise constant derivative. (\textcolor{red}{Use regularized derivative?})

@@

	Here we demonstrate that, with decreasing time step $\Delta t$ and $\alpha=0$, the estimated rates converge to the true rates if there is no ambiguity in the system. Let there be two species $A$ and $B$ which can unimolecularly convert into each other with macroscopic rates of $k_1=4.0$ and $k_2=0.5$, respectively. The initial amount of particles is set to $70$ of type $A$ and $0$ of type $B$. The to this system corresponding law of mass action solution and a Gillespie realization are  depicted in Figure \ref{ex1:concentration-curves}.

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/aux_ex1_concentration_curves}
		\end{center}
		\caption{The law of mass action solution and a Gillespie realization of the example given in Section \ref{sec:ex1}.}
		\label{ex1:concentration-curves}
	\end{figure}

	For each $\Delta t$, eight different realizations with identical starting conditions are averaged into a single trajectory, of which then the rates for reactions $A\to B$ and $B\to A$ are estimated. This procedure was repeated $30$ times. In Figure \ref{ex1:rates-convergence}, the average estimates with their standard deviations are shown. One can see that, with decreasing $\Delta t$, the estimates become better and the rates can be recovered. Further it can be observed that in this case a time step of $\Delta t\approx 10^{-2}$ suffices.

	\begin{figure}
		\begin{center}
			\includegraphics[width=.5\textwidth]{./figures_tex/aux_rates_convergence}
		\end{center}
		\caption{Convergence of estimated rates with decreasing $\Delta t$.}
		\label{ex1:rates-convergence}
	\end{figure}

@@

	The problem now is to find a suitable $\alpha$ such that a parsimonious solution is yielded. To this end, cross-validation is applied. The trajectories that are used as training and test data set are depicted in Figure \ref{ex2:test-train-concentration-curves}. \textcolor{red}{Perhaps use several test trajectories with different initial conditions to get something like a confidence interval?} As an indicator to the estimation error one can use the scoring function
	\begin{align}
		f_\alpha = \frac{1}{T}\|\mathbf{\dot{X}}_\text{test} - \Theta (\textbf{X}_\text{test}) \Xi_\text{train}\|_F^2,
		\label{ex2:score-function}
	\end{align}
	where the best possible score for some $f_\alpha = 0$ and larger values indicate worse models.
	For a selection of hyperparameters $\alpha\in [0, 100]$, the model parameters $\Xi$ are estimated and the scoring function (\ref{ex2:score-function}) evaluated, yielding a dependency of the score on $\alpha$ as depicted in Figure \ref{ex2:score-graph}.

@@

	One can see that the minimum score is achieved at
	\begin{align}
		\hat{\alpha} = \underset{\alpha\in[0,100]}{\arg\min}f_\alpha\approx 10.
	\end{align}
	When evaluating the minimization problem (\ref{method:minimizationproblem}) with $\hat{\alpha}$ as hyperparameter for the regularizer and applying a cut-off of $10^{-10}$ to the yielded parameters, one obtains the simpler set of reactions $A\xrightarrow{\hat{k}_1}D$, $D\xrightarrow{\hat{k}_3} B$, and $A+B\xrightarrow{\hat{k}_4}C$ with rates $\hat{k}_1 \approx 3.5\cdot 10^{-3}$, $\hat{k}_3\approx 9.1\cdot 10^{-3}$, and $\hat{k}_4\approx 8.7\cdot 10^{-5}$, respectively.
